---
title             : "The title"
shorttitle        : "Title"

author: 
  - name          : "First Author"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
  - name          : "Ernst-August Doelle"
    affiliation   : "1,2"

affiliation:
  - id            : "1"
    institution   : "Wilhelm-Wundt-University"
  - id            : "2"
    institution   : "Konstanz Business School"

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

  Enter author note here.

abstract: |

  @Hamlin2007's study is very cool, but some studies are just as cool [@Hamlin2011].
  
  <!-- https://tinyurl.com/ybremelq -->
  
keywords          : "keywords"
wordcount         : "X"

bibliography      : ["References.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
library(tidyverse)
library(knitr)
knitr::opts_chunk$set(fig.width=8, fig.height=5,
               echo=FALSE, 
               warning=FALSE, message=FALSE, 
               cache=FALSE)
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(42)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```



# Methods


We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.


# Results

Gear 3 and 5 have higher horsepower.

```{r fig1, fig.cap = "\\label{fig:fig1}Horsepower by number of gears."}
data <- mtcars
data <- data %>%
  group_by(gear) %>%
  summarize(mean_hp = mean(hp))
data$gear <- as.factor(data$gear)
ggplot(data, aes(x=gear, y=mean_hp)) +
  geom_col(position=position_dodge(0.9)) + 
  theme_apa()
```


```{r fig2, results = 'asis',fig.cap = "\\label{fig:fig2}Horsepower by number of gears, table."}
apa_table(data)
```

# Discussion


\newpage

# References
```{r create_r-references}
r_refs(file = "r-references.bib")
```

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id = "refs"></div>
\endgroup
