---
title: "Tidyverse and documentation"
subtitle: ""
author: "Francis L. Yuen"
date: "`r Sys.Date()`"
output:
  tufte::tufte_html: default
  tufte::tufte_handout:
    citation_package: natbib
    latex_engine: xelatex
  tufte::tufte_book:
    citation_package: natbib
    latex_engine: xelatex
bibliography: skeleton.bib
link-citations: yes
---

```{r setup, include=FALSE}
library(apa)
library(knitr)
library(tufte)
library(tidyverse)
library(here)
library(janitor)
opts_chunk$set(echo=TRUE, 
               warning=FALSE, message=FALSE, 
               cache=FALSE)
```

Learning goals:

1. Writing readable scripts using the pipe function
2. Properly document your script
3. Practice cleaning, tidying, analyzing, and visualizing data

# Writing readable scripts

In the last workshop, we wrote a series of codes that analyzed a mock dataset. Open up that R script. It should look something like this:
```{r include = FALSE}
data <- read_csv(here("R Workshops/3. Tidyverse and documentation/data", "sample_data_1.csv"))
data <- clean_names(data)
```
```{marginfigure}
Remember to check that your raw data is stored in the right place so that `here()` can find it. Adjust the directory to help `here()` locate your raw data if necessary.
```


```{r eval=FALSE}
library(tidyverse)
library(here)
data <- read_csv(here("data", "sample_data_1.csv"))
data <- clean_names(data)
data_tidy <- pivot_longer(data = data,
                     cols = c(trial1:trial10),
                     names_to = "trial_number",
                     names_prefix = "trial",
                     values_to = "looking_time")
data_tidy <- mutate(.data = data_tidy,
                    corrected_looking_time = looking_time + 3.4,
                    log_looking_time = log(corrected_looking_time))
data_tidy <- filter(.data = data_tidy,
                    corrected_looking_time <= 30.00)
data_tidy <- group_by(.data = data_tidy,
                      hypothesis)
lt_by_choice <- summarize(.data = data_tidy,
                          mean = mean(log_looking_time),
                          sd = sd(log_looking_time),
                          n = length(log_looking_time),
                          SEM = sd/sqrt(n))
```


Recall that one of the problems here is that you have to execute several separate lines, each one updating what 'data_tidy' is before finally arriving at your results. We can shorten the code by using a function called the pipe, which is written like this: `%>%`

What the pipe does is **take the object that comes before the pipe, and pass it on as the first argument of the function that comes after the pipe**. To illustrate, let's use the pipe to rewrite the first part of our code and store it as `data_tidy2` so we can compare the results:

```{r eval = FALSE}
data_tidy2 <- data %>%
  pivot_longer(cols = c(trial1:trial10),
               names_to = "trial_number",
               names_prefix = "trial",
               values_to = "looking_time")
```

